A novel approach to chemical tagging in lower-dimensional latent space

### 2018-12-19 article/ms.tex
foo

As mentioned previously, the model as described is indeterminate in that
there is no unique solution for the factor loads $\factorloads$ and scores
$\factorscores$. These quantities can only be determined up until 
orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
can be obtained by the expectation-maximization algorithm \citep{EM}.
%The model described by Equation~\ref{eq:generative-model} is indeterminate:
%there is no unique solution for the factor loads $\factorloads$ and scores
%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.

@@

We use the expectation-maximization algorithm to estimate the model parameters.
With each expectation step we evaluate the log likelihood given the model 
parameters $\vec\Psi$, and we re-calculate the responsibility 
matrix $\vec\tau$ whose entries are the posterior probability
that the $\numdata$th data point is associated to the $\numcomponents$th component,
given the data $\vecdata$ and the current estimate of the 
parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
\end{equation}

@@

\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeated
the expectation-maximization algorithm for up to 1,000 steps or until the log-likelihood
improved by less than $10^{-5}$ between successive iterations. Although our implementation
allows for a positive regularisation term to be added along the diagonal of the covariance
matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
experiments performed here.

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, and the
Bayesian Information Criterion \citep{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
\end{equation}

These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value is found at $\NumLatentFactors = 5$
and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_{true} = 5$,
$\NumComponents_{true} = 20$).

We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
In this figure we have over-plotted 100000 samples drawn from the model
with $J = 5$ and $K = 21$
-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:exp1-latent-space} where the colours indicate the
most likely component that the data point is associated with.

>>>>

As mentioned previously, the model as described is indeterminate in that
there is no unique solution for the factor loads $\factorloads$ and scores
$\factorscores$. These quantities can only be determined up until 
orthogonality. However, accurate estimates of the parameter vector $\vec\Psi$
can be obtained by the expectation-maximization algorithm \citep{Dempster:1977}.
%The model described by Equation~\ref{eq:generative-model} is indeterminate:
%there is no unique solution for the factor loads $\factorloads$ and scores
%$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.

@@

We use the expectation-maximization algorithm to estimate the model parameters
\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
given the model parameters $\vec\Psi$, and we re-calculate the responsibility 
matrix $\vec\tau$ whose entries are the posterior probability that the 
$\numdata$th data point is associated to the $\numcomponents$th component, given 
the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \eye\specificvariance)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\specificvariance)} \quad .
\end{equation}

@@

\noindent{}where $\odot$ denotes is the entry-wise (Hadamard) product. We repeat
the expectation-maximization cycle for up to 10,000 steps or until the log-likelihood
improved by less than $10^{-5}$ between successive iterations. Although our implementation
allows for a positive regularisation term to be added along the diagonal of the covariance
matrices in latent space $\scorecovs$, we set this regularisation term to zero for all
experiments performed here.

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_\textrm{true}$ and $\NumComponents_\textrm{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, and the
Bayesian Information Criterion \citep[BIC][]{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
\end{equation} 
\noindent{}where $Q$ is the number of model parameters:
\begin{equation}
	%Q = \NumComponents + \NumDimensions - 1 + \NumLatentFactors(\NumDimensions + \NumComponents) + \frac{1}{2}\NumComponents\NumLatentFactors(\NumLatentFactors+1) - \NumLatentFactors^2 \quad .
	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1
\end{equation}

These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the log likelihood increases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value is found at $\NumLatentFactors = 5$
and $\NumComponents = 21$, close to the true values ($\NumLatentFactors_\textrm{true} = 5$,
$\NumComponents_\textrm{true} = 20$).

We show a corner plot of the data in Figure~\ref{fig:exp1-data}.
In this figure we have over-plotted 100000 samples drawn from the model
with $J = 5$ and $K = 21$
-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:exp1-latent-space} where the colours indicate the
most likely component that the data point is associated with.
